---
title: "Demonstration of tidyext"
author: "Michel Clark"
date: <span style="font-style:normal;font-family:'Open Sans'">`r Sys.Date()`</span>
output:
  html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable

vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Demonstration of tidyext}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, collapse = TRUE,
                      comment=NA, R.options=list(width=220),   # code 
                      dev.args=list(bg = 'transparent'), dev='svglite',                                 # viz
                      fig.align='center', out.width='75%', fig.asp=.75,                 
                      cache.rebuild=F, cache=F)                                                         # cache
```


# Getting started with tidyext

## Data Summaries

To begin, we can load up the <span class="pack">tidyverse</span> and this package. I'll also create some data that will be useful for demonstration.

```{r packages}
library(tidyverse); library(tidyext)
set.seed(8675309)
df1 <- tibble(
  g1 = factor(sample(1:2, 50, replace = TRUE), labels=c('a','b')),
  g2 = sample(1:4, 50, replace = TRUE),
  a = rnorm(50),
  b = rpois(50, 10),
  c = sample(letters, 50, replace=TRUE),
  d = sample(c(T,F), 50, replace=TRUE)
)

df_miss = df1
df_miss[sample(1:nrow(df1), 10), sample(1:ncol(df1), 3)] = NA
```

We can start by getting a quick numerical summary for a single column.  As the name suggests, this will only work with numeric data.

```{r num_summary}
num_summary(mtcars$mpg)

num_summary(df_miss$a, extra = T)
```



Note that the result's class is a <span class="objclass">data.frame</span>, which makes it easy to work with.

```{r class_num_sum}
x = num_summary(mtcars$mpg)
glimpse(x)

mtcars %>% 
  map_dfr(num_summary, .id = 'Variable')
```

There are also functions for summarizing missingness.

```{r missing}
sum_NA(df_miss$a)

sum_blank(c(letters, '', '   '))

sum_NaN(c(1, NaN, 2))
```


When dealing with a data frame of mixed types we can use the <span class="func">describe_*</span> functions.  

```{r describe}
describe_all(df1)

describe_all_cat(df1)

describe_all_num(df1, digits = 1)
```


Note how the categorical data result is just as ready for visualization as the numeric, as it can be filtered by the `Variable` column.  It also has an option to deal with <span class="objclass">NA</span> and some other stuff.

```{r describe_cat, out.width='50%'}
describe_all_cat(df_miss, include_NAcat = TRUE, sort_by_freq = TRUE) %>% 
  filter(Variable == 'g1') %>% 
  ggplot(aes(x=Group, y=`%`)) +
  geom_col(width = .25)
```


Typically during data processing, we are performing grouped operations.  As such there is a corresponding <span class="func">num_by</span> and <span class="func">cat_by</span> to provide the same information by some grouping variable.  This basically is saving you from doing `group_by %>% summarize()` and creating variables for all these values.  It can also take a set of variables to summarize using <span class="func">vars</span>.

```{r num_by}
df_miss %>% 
  num_by(a, group_var = g2)

df_miss %>% 
  num_by(vars(a, b), group_var = g2)
```

For categorical variables summarized by group, you can select whether the resulting percentage is irrespective of the grouping.

```{r cat_by}
df1 %>% 
  cat_by(d, 
         group_var = g1, 
         perc_by_group = TRUE)

df1 %>% 
  cat_by(d, 
         group_var = g1, 
         perc_by_group = FALSE, 
         sort_by_group = FALSE)
```



## Data Processing

In addition there are some functions for data processing.  We can start with the simple one-hot encoding function.

```{r onehot}
onehot(iris) %>% 
  slice(c(1:2, 51:52, 101:102))
```

It can do it sparsely.

```{r onehot2}
iris %>% 
  slice(c(1:2, 51:52, 101:102)) %>% 
  onehot(sparse = TRUE)
```

Chose a specific variable, whether you want to keep the others, and how to deal with NA.

```{r onehot3}
df_miss %>%
  onehot(var = c('g1', 'g2'), nas = 'na.omit', keep.original = FALSE) %>%
  head()
```


With <span class="func">create_prediction_data</span>, we can quickly create data for use with <span class="func">predict</span> after a model.  By default it will put numeric variables at their mean, and categorical variables at their most common category.

```{r create_prediction_data}
create_prediction_data(iris)

create_prediction_data(iris, num = function(x) quantile(x, p=.25))
```

We can also supply specific values.


```{r create_prediction_data2}
cd = data.frame(cyl=4, hp=100)
create_prediction_data(mtcars, conditional_data = cd)
```

For modeling purposes, we often want to center or scale the data, take logs etc.  The <span class="func">pre_process</span> function will standardize numeric data by default.

```{r preprocess}
pre_process(df1)
```

Other options are to simply center the data (`scale_by = 0`), start some variables at zero (e.g. time indicators), log some variables (with chosen base), and scale some to range from zero to one.  

```{r preprocess2}
pre_process(mtcars, 
            scale_by = 0, 
            log_vars = vars(mpg, wt), 
            zero_start = vars(cyl), 
            zero_one = vars(hp, starts_with('d'))) %>% 
  describe_all_num()
```

Note that center/standardizing is done to any numeric variables *not* chosen for `log`, `zero_start`, and `zero_one`.  

Here's a specific function you will probably never need, but will be glad to have if you do.  Some data columns have multiple entries for each observation/cell.  While it's understandable why someone would do this, it's not very good practice. This will split out the entries, or any particular combination of them, into their own indicator column.

```{r combn, message=TRUE}
d = data.frame(id = 1:4,
               labs = c('A-B', 'B-C-D-E', 'A-E', 'D-E'))
combn_2_col(data=d, var='labs', max_m=2, sep = '-', collapse = ':', toInteger = T)

combn_2_col(data=d, var='labs', max_m=2, sparse = T)
```

In addition, there is a function that makes <span class="pack">tidyr's</span> <span class="func">spread</span> work like it should when you don't have unique identifiers.  I spent a lot of time coming up with the name for this one.

```{r spread2, error=TRUE}
# initial example from spread
stocks_init <- data.frame(
  time = as.Date('2009-01-01') + 0:9,
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)

# a very common situation
stocks <- data.frame(
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)


stocksm_init <- stocks_init %>% 
  gather(stock, price, -time)
head(stocksm_init)

stocksm_init %>% 
  spread(stock, price)  %>% 
  head()                                 # works fine

# no time
stocksm <- stocks %>% 
  gather(stock, price)
head(stocksm)

stocksm %>% 
  spread(stock, price)         # annoying

stocksm %>% 
  spread2(stock, price)        # works fine
```

However, with unbalanced data the result can be interpreted in different ways, so you'll have a choice to make.  

```{r spread2compact}
stocksm$price[sample(1:nrow(stocksm), 5)] = NA
stocksm %>% 
  spread2(stock, price)

stocksm %>% 
  spread2(stock, price, compact = FALSE)
```

I can speak from experience that having longitudinal data in the `compact = FALSE` format is about the worst way you could keep such data.

Use <span class="func">gather_multi</span> to extend `tidyr::gather` to deal with multiple sets of variables.

```{r gather_multi}
demo_data_wide = data.frame(id = 1:10,
                            X = matrix(rnorm(40), ncol = 4),
                            Y = matrix(sample(0:1, 40, replace = T), ncol = 4),
                            Z = matrix(rpois(40, 5), ncol = 4))
head(demo_data_wide)

gather_multi(demo_data_wide,
             key = wave,
             values = vars(X, Y, Z),
             varlist = c(vars(starts_with('X')),
                         vars(starts_with('Y')),
                         vars(starts_with('Z'))),
             -id)
```

Note that this only makes sense with balanced data, though it will work if NAs are present.  Also, you may want to change the key in some fashion.

<br>
<br>
<br>
<br>
<br>
<br>
