---
title: "Demonstration of tidyext"
author: "Michel Clark"
date: <span style="font-style:normal;font-family:'Open Sans'">`r Sys.Date()`</span>
output:
  html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable

vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Demonstration of tidyext}
  %\VignetteEncoding{UTF-8}
# params:
#   EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Demonstration of tidyext}
%\VignetteEncoding{UTF-8}
-->

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, collapse = TRUE,
                      comment=NA, R.options=list(width=220),   # code 
                      dev.args=list(bg = 'transparent'), dev='svglite',                                 # viz
                      fig.align='center', out.width='75%', fig.asp=.75,                 
                      cache.rebuild=F, cache=F)                                                         # cache
```


# Getting started with tidyext

## Data Summaries

To begin, we can load up the <span class="pack">tidyverse</span> and this package. I'll also create some data that will be useful for demonstration.

```{r packages}
library(tidyverse); library(tidyext)
set.seed(8675309)
df1 <- tibble(
  g1 = factor(sample(1:2, 50, replace = TRUE), labels=c('a','b')),
  g2 = sample(1:4, 50, replace = TRUE),
  a = rnorm(50),
  b = rpois(50, 10),
  c = sample(letters, 50, replace=TRUE),
  d = sample(c(T,F), 50, replace=TRUE)
)

df_miss = df1
df_miss[sample(1:nrow(df1), 10), sample(1:ncol(df1), 3)] = NA

```

We can start by getting a quick numerical summary for a single column.  As the name suggests, this will only work with numeric data.

```{r num_summary}
num_summary(mtcars$mpg)
num_summary(df_miss$a, extra = T)
```



Note that the result's class is a <span class="objclass">data.frame</span>, which makes it easy to work with.

```{r class_num_sum}
x = num_summary(mtcars$mpg)
glimpse(x)
mtcars %>% 
  map(num_summary) %>% 
  bind_rows()
```

There are also functions for summarizing missingness.

```{r missing}
sum_NA(df_miss$a)
sum_blank(c(letters, '', ''))
sum_NaN(c(1, NaN, 2))
```


When dealing with a data frame of mixed types we can use the <span class="func">describe_*</span> functions.  

```{r describe}
describe_all(df1)
describe_all_cat(df1)
describe_all_num(df1, digits = 1)
```


Note how the categorical data result is just as ready for visualization as the numeric, as it can be filtered by the `Variable` column.  It also has an option to deal with <span class="objclass">NA</span> and some other stuff.

```{r describe_cat}
describe_all_cat(df_miss, include_NAcat = TRUE, sort_by_freq = TRUE) %>% 
  filter(Variable == 'd') %>% 
  ggplot(aes(x=Group, y=`%`)) +
  geom_col(width = .25)
```


Typically during data processing, we are performing grouped operations.  As such there is a corresponding num_by and cat_by to provide the same information by some grouping variable.  This basically is saving you from doing `group_by %>% summarize()` and creating variables for all these values.  It can also take a set of variables using <span class="func">vars</span>

```{r num_by}
df_miss %>% 
  num_by(a, group_var = g2)

df_miss %>% 
  num_by(vars(a, b), group_var = g2)
```

For categorical variables summarized by group, you can select whether the resulting percentage is irrespective of the grouping.

```{r cat_by}
df1 %>% 
  cat_by(d, 
         group_var = g1, 
         perc_by_group = TRUE)

df1 %>% 
  cat_by(d, 
         group_var = g1, 
         perc_by_group = FALSE, 
         sort_by_group = FALSE)
```



## Data Processing

In addition there are some functions for data processing.  We can start with the simple one-hot encoding function.

```{r onehot}
onehot(iris) %>% 
  slice(c(1:2, 51:52, 101:102))
```


We can quickly create data for use with <span class="func">predict</span>, and a function for pre-processing data for use with a model.  By default it will put numeric variables at their mean, and categorical variables at their most common category.

```{r create_prediction_data}
create_prediction_data(iris)
create_prediction_data(iris, num = function(x) quantile(x, p=.25))
```

We can also supply specific values.


```{r create_prediction_data2}
cd = data.frame(cyl=4, hp=100)
create_prediction_data(mtcars, conditional_data = cd)
```

For modeling purposes, we often want to center or scale the data, take logs etc.  The pre_process function will standardize numeric data by default.

```{r preprocess}
pre_process(df1)
```

Other options are to simply center the data (`scale_by = 0`), start some variables at zero, log some variables (with chosen base), and scale some to range from zero to one.  

```{r preprocess2}
pre_process(mtcars, 
            scale_by = 0, 
            log_vars = vars(mpg, wt), 
            zero_start = vars(cyl), 
            zero_one = vars(hp, starts_with('d'))) %>% 
  describe_all_num()
```

Note that center/standardizing is done to any variables *not* chosen for log, zero_start, and zero_one.  

Here's a specific function you probably will never need, but will be glad if you do.  Some data columns have multiple values.  While understandable, it's not very good practice.

```{r combn}
d = data.frame(id = 1:4,
               labs = c('A/B', 'B/C/D/E', 'A/E', 'D/E'))
combn_2_col(data=d, var='labs', max_m=3)
```

And finally, there is a function that makes <span class="pack">tidyr's</span> <span class="func">spread</span> work like it should when you don't have unique identifiers.  

```{r spread2, error=TRUE}
# initial example from spread
stocks_init <- data.frame(
  time = as.Date('2009-01-01') + 0:9,
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)

# a very common situation
stocks <- data.frame(
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)


stocksm_init <- stocks_init %>% gather(stock, price, -time)

stocksm_init %>% 
  spread(stock, price)  %>% 
  head()                                 # works fine


stocksm <- stocks %>% gather(stock, price)
stocksm %>% spread(stock, price)         # annoying
stocksm %>% spread2(stock, price)        # works fine
```

However, with unbalanced data the result can be interpreted in different ways, so you'll have a choice to make.

```{r spread2compact}
stocksm$price[sample(1:nrow(stocksm), 5)] = NA
stocksm %>% spread2(stock, price)
stocksm %>% spread2(stock, price, compact = FALSE)
```


<br>
<br>
<br>
<br>
<br>
<br>
